Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-04-28 01-introduction.qmd
review introduction and antecedents

Meritocracy posits that inequality is an inherent feature of societies but can be legitimized through fairer principles, such as effort and talent [@young_rise_1962; @davis_principles_2001]. Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality since these beliefs are associated with attitudes that maintain the legitimization of social differences [@batruch_belief_2023]. In a context like Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], these beliefs can play a crucial role in the acceptance of social inequalities. Incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified. Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].

>>>>

Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001]. Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that  legitimize status differences [@batruch_belief_2023]. In a context like Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], these beliefs can play a crucial role in the acceptance of social inequalities. Incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified. Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].

### 2025-04-28 02-antecedents.qmd
review introduction and antecedents

## Justification of market inequality

@@

Empirical studies have developed different strategies to capture and operationalize market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al. [-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skill. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, or pensions. For example, Knesebeck et al. (2016) and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Some research, like that of Lindh [-@lindh_public_2015] and Svallfors [-@svallfors_political_2007], uses these indicators in combination to measure broader preferences for market-based distribution in welfare systems. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These different operationalizations all aim to capture the extent to which individuals accept inequality when it is framed as a reflection of market outcomes.

@@

@@

Importantly, perceptions of inequality have been linked to attitudes about market justice---the belief that goods and services should be allocated according to individual contribution, competition, or willingness to pay [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

@@

Meritocracy constitutes a central ideological framework for legitimizing social inequality. Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions. As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality. However, over time, meritocracy has been widely adopted in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020]. From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality. People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_percepcion_2019; @batruch_belief_2023]. Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust. This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.

@@

## This study (hypotheses)

>>>>

## The justification of market inequality

@@

Empirical studies have developed different strategies to capture and operationalize market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al. [-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skills. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions. For example, Knesebeck et al. (2016) and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Some research, like that of Lindh [-@lindh_public_2015] and Svallfors [-@svallfors_political_2007], uses these indicators in combination to measure broader preferences for market-based distribution in welfare systems. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These different operationalizations all aim to capture the extent to which individuals accept inequality when it is framed as a reflection of market outcomes.

@@

Perceptions of inequality have been associated to attitudes about market justice[@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

@@

@@

Meritocracy constitutes a central ideological framework for legitimizing different types od social inequality, for instance through market justice beliefs. Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions. As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality. However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020]. From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality. People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_percepcion_2019; @batruch_belief_2023]. Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust. This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.

@@

## This study

Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their beliefs in meritocracy. Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile.

First, consistent with previous findings, we expect that a lower perception of economic inequality will be associated with stronger market justice preferences. When individuals perceive smaller economic gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing acceptance of outcomes based on competition and ability to pay. Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution. This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force shaping public attitudes toward inequality and justice.

Second, stronger meritocratic beliefs are expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit. This aligns with the idea that meritocratic beliefs serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes. Specifically, we argue that the legitimizing effect of meritocratic beliefs on market justice preferences is moderated by perceptions of economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger. However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.

Additionally, this study examines whether these relationships vary over time in response to significant social and political events. In particular, we explore whether the political outburst of 2019 and the subsequent constitutional processes in Chile have altered how perceptions of inequality and meritocracy shape market justice attitudes. We expect that after these critical events, perceptions of inequality may have a stronger negative effect on market justice preferences, reflecting increased societal questioning of market-based allocation mechanisms.

Based on these arguments, we propose the following hypotheses:

H1: Lower perceived economic inequality is associated with stronger market justice preferences.

H2: Stronger meritocratic beliefs are associated with stronger market justice preferences.

H3: The positive association between meritocratic beliefs and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.

H4: The effects described in H1–H3 are attenuated after major social mobilizations (2019–2022), reflecting increased critical views of the market's role in allocating social goods.
